---
title: "mini-DREAM Exercise & Evaluation 1"
author: "James Eddy"
date: "7/6/2017"
output: html_notebook
---

# About this exercise

It's tough for us to guage how well everyone is following along during the mini-DREAM videoconferences. While we've tried to keep the workload for mini-DREAM minimal outside of the meetings — and have structured activities so that a firm understanding of the programming and stats isn't required — we'd still like to get a sense for what you *are* learning!

This exercise is designed to see what concepts you can recall from previous activities. For each part, you'll need to fill in blanks that look like this "_____" in order to run the chunk without getting an error. Note that the number of blanks *does not* match the number of missing letters in the code. This is "open book" — you can switch between modules in RStudio, or look back at past activities on Synapse to view those notebooks as needed.

If you get stuck or can't figure out a step, use the WebEx chat to ask questions. This exercise isn't about testing you, but is a way for us to see what we might need to focus on extra.

---

# Load some data

You've seen a few different ways to load data into R. Fix the commands in the chunks below to load some of the METABRIC data that we've been using for activities. 

## 1. Load a tab-delimited file

We've mostly been working with an "RData" file to conveniently load both clinical and expression data. However, in **Activity 1**, we also used a different command to load data directly from the original tab-delimited text (`.txt`) file. Use that command here to load the clinical data table.

```{r}
clinical_table_file <- "/home/shared/data/metabric_split/activity_clinical.txt"
activity_clinical_df <- read._____(clinical_table_file, delim = _____)
```

## 2. Load an RData file

Now use the command to load both the clinical and expression tables from the `RData` file.

```{r}
activity_dataset_file <- "/home/shared/data/metabric_split/activity_data.RData"
_____(activity_dataset_file)
```

---

# Inspect the data

We've also covered a number of approaches for checking the structure and type of data, once it's been loaded into R.

## 3. Preview rows in a data frame

Print the first 6 rows of `activity_clinical_df` using a certain function (i.e., not by subsetting the data). 

```{r}
_____(activity_clinical_df)
```

## 4. Check the size of the data

Check how many genes are in the expression table (**hint:** think about whether genes are listed in rows or columns of the table and use the corresponding function to count).

```{r}
_____(activity_expression_df)
```

## 5. Check variable type

Each of the patients in the data have been assigned a value based on their "intrinsic clustering membership"" determined by METABRIC investigators. Print the type of the corresponding variable in `activity_clinical_df`.

The name of the variable might not be obvious — use another function first to view all column headers in the data frame.

```{r}
_____(activity_clinical_df)
```

Now check the type of that variable.

```{r}
_____(activity_clinical_df$_____)
```

---

> FAST FORWARD...

(Data cleaning, formatting, arranging, and other various munging tasks are a bit of a learned art. Because you haven't had a lot of direct practice with these operations, we'll skip over this step and load a downstream version of the data that is ready for analysis.)

Just in case you weren't able to load the data above, run this chunk to get the variables you'll need for subsequent steps.

```{r}
uncensored_clin_df <- readRDS("/home/shared/data/metabric_split/activity_uncensored_clin_df.rds")
uncensored_expr_mat <- readRDS("/home/shared/data/metabric_split/activity_uncensored_expr_mat.rds")
```

---

# Make a plot

## 6. Collect expression values for a gene

Print the row in the `uncensored_expr_mat` matrix where we can find expression values for the gene *of your choice*. Replace the text (<fill-in-gene-name-here>) with the gene name you want to use (e.g., AURKA, EGFR, AVP, etc.).

```{r}
gene_row <- which(rownames(uncensored_expr_mat) _____ "<fill-in-gene-name-here>")
gene_row
```

Print the expression value of gene for the 142nd patient in the data.

```{r}
gene_uncensored <- as.vector(uncensored_expr_mat[gene_row, ])
gene_uncensored_____142_____
```

## 7. Plot survival time vs. gene expression

Create a scatterplot that shows survival time (y-axis) as a function of gene expression (x-axis).

```{r}
_____(uncensored_clin_df$T _____ gene_uncensored)
```

---

# Build a model

## 8. Fit a linear model

Train a linear regression model for the relationship between survival time and gene expression and save it to the variable `gene_model`.

```{r}
gene_model <- _____(uncensored_clin_df$T ~ gene_uncensored)
```

## 9. Inspect model results

Use a function to summarize some of the key details of the model (residuals, coefficients, R-squared, etc.).

```{r}
_____(gene_model)
```

---

## Save & upload your notebook

When you're done, save this file and run the chunk below, which will upload the Rmd to Synapse. Nothing gets submitted to the mini-DREAM challenge for this exercise.

```{r}
library(synapseClient)
synapseLogin(rememberMe = TRUE)
notebook_filename <- paste(Sys.getenv("USER"), "exercise1.Rmd", sep = "_")
exercise1_notebook <- synStore(File(path = notebook_filename, parentId = "syn10163686"))
```
