---
title: "Module 1 - Intro to Programming with R"
author: "Alexendar Perez & James Eddy (credit to Alvaro Gonzalez for code development)"
date: "June 15, 2017"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# About this activity

You won't need to write any new code in this file. You don't really even need to understand what the code is doing or how. We could spend an entire summer teaching you the basics of programming in R, so it would be unreasonable to expect you to be self-sufficient in R after a single hour-long activity.

Instead, we want you to read through the notebook and run each chunk like you did in **Module 0**. Look at each output, and think about what's happening. Our hope is that you come away from this activity thinking "I might not know how to code by myself in R, but I see that it can do some pretty useful things for me!" 

Throughout the rest of the course, the approach will be similar — follow along and observe that you're *using* R to complete these data-centric tasks, but don't worry too much if you don't really get *how* it's working. Any concepts that we want you to understand and be able to use on your own, we'll let you know (and you'll probably see those concepts over and over throughout the challenge modules).

---

# Nuts and bolts of R

When we talk about "data" in R, we usually mean *collections* of information that vary in size, shape, and complexity. One of the most common forms of data we encounter is a table, which is similar in a lot of ways to spreadsheets you've seen in Excel. Before we start working with example data in R, it's worth taking some time to go over some of the core functionality that R provides.

## Commands in R

The basic flow for interacting with R is to write one or more *commands*. These commands provide the instructions that tell R what to do. For example, I can use a command that combines the keyword `print` with some message like "Hello world!" to instruct R to output that message.

```{r}
# this is a comment -- R will ignore lines beginning with '#'
print("Hello world!")
```

Anytime you see an R command in the form `keyword()`, that represents a **function** — this tells R to perform that function on whatever *inputs* are inside the parentheses. Usually, functions will be named like verbs, which is another useful clue that they carry out some action:

```{r}
# 'round' 1.8 to the nearest whole number
round(1.8)
```

Unfortunately, this isn't always true... When you see a function that's named as a noun, you can typically think of this as "give me the `keyword` for this input".

```{r}
# show the 'ceiling' of 1.8, which is the result of rounding up to the nearest whole number
ceiling(1.8)
```

Whenever you run a command that is just a value or a name, you're telling R to show that value:

```{r}
# print the value 1
1
```

```{r}
# print "hello"
"hello"
```


## R as a calculator

One of the simplest ways to use R is to perform basic arithmetic. Step through the following chunks to see how R can do calculations for you.

Add or subtract numbers using the '+' and '-' symbols:

```{r}
# addition
1 + 2
```

```{r}
# subtraction
3 - 2
```

Multiply or divide numbers using the '*' and '/' symbols

```{r}
# multiplication
2 * 3
```

```{r}
# division
6 / 3
```

Use '^' to raise a number to some power

```{r}
# exponentiation
2 ^ 3
```

Note that for some common operations, R also provides some convenient short-cut functions. For example, you can use `sqrt()` to get the square root of a number:

```{r}
sqrt(4)
```

Which is the same as raising that number to the power of 1/2:

```{r}
4 ^ (1/2)
```

### Order of operations

When calculating results, R uses the same order of operations you learned in algebra:

```{r}
# multiplication or division before addition or subtraction
1 + 2 * 3
```

```{r}
# operations within parentheses first
(1 + 2) * 3
```

```{r}
# otherwise, left to right
1 * 2 / 3
```

## Using variables

As you probably already guessed, R is more than just a calculator. Like many other programming languages, R has the ability to link values to names or *variables*. By assigning a value to a variable, we can more easily use that value in other operations. In R, we assign values to a variable using the symbols `<-` (like a left arrow):

```{r}
# assign the value 100 to the variable x
x <- 100
```

`x` is a variable that now holds the value `100`.

Typing the name of the variable on the command line prints out the associated value:
```{r}
x
```

Now we can perform operations on the value connected to `x` (without changing that value):

```{r}
x + 25
```

```{r}
x / 5
```

The above example probably isn't very impressive — it's not too hard to retype `100`. However, variables can be used to save more complicated values as well.

```{r}
# Avogadro's constant
L <- 6.022140857 * 10^23
L
```

```{r}
L * 10
```

Another advantage of using variables is that we can perform a series of operations and only show the final value when we're done.

```{r}
x <- 1
y <- x + 3
z <- y * 4
z
```

## Data types

All variables belong to a data type. Data types are called classes in R. A class defines the characteristics of the data stored in the variable. R provides a number of atomic or elementary data types. Asking R to describe a value's type is done with the **class** command.

Below are some elementary data types supported by R. You don't need to worry about what exactly these types mean or how they're used right now — run the code chunks to get a feel for what each type looks like.

1. character

```{r}
# character values are surrounded by quotes
x_char <- "this is a character value"
class(x_char)
```

```{r}
print(x_char)
```

2. numeric or "double" (real numbers)

```{r}
# numeric values are any numbers that might have decimals
x_num <- 1.4
class(x_num)
```

3. integer

```{r}
# integer values are whole numbers, indicated by an "L"
x_int <- 1L
class(x_int)
```

```{r}
print(x_int)
```

4. logical (boolean)

```{r}
# logical values can either be TRUE or FALSE
x_logical <- TRUE
class(x_logical)
```

```{r}
print(x_logical)
```

You'll see each of these data types again in future modules, and you'll occasionally need to pay attention to the types of variables before trying to perform calculations. We'll show you additional ways to check data types and also convert between types, if needed.

## Vectors

R starts to get more interesting (and useful) when we want to work with *sets* of values. Vectors are collections of objects of the same data type. For example, a vector of numerics is a collection of real numbers.

Vectors can be created using the **c** function. A fundamental attribute of vectors is the length.

```{r}
# numeric vector
x <- c(10, 20)
length(x)
```

```{r}
# character vector
x <- c("A", "B", "C")
x
```

```{r}
# logical vector
x <- c(TRUE, FALSE, TRUE)
x
```

A shortcut for creating vectors of integers is the `:` symbol, which tells R that you want a series of all integers between two numbers.

```{r}
# integer vector
x <- 1:5
x
```

You can also reverse the numbers to create a vector with integers in descending rather than ascending order.

```{r}
x <- 5:1
x
```

More generally, the ```seq``` function can be used to create integer vectors in any given range. Let's use R's powerful help command:
```{r}
?seq
```

```{r}
# create a sequence from 1 to 3, incrementing by 1
seq(1, 3)
```

```{r}
# create a sequence from 1 to 20, incrementing by 5
seq(1, 20, 5)
```

## And more...

There are many other concepts we could cover about the basics of R (and even more, for advanced R topics!). Rather than try to go over all of these concepts today, we'll introduce and explain new things as they come up in each module.

---

# Working with data in R

Now that you've seen a little bit of what it means to use R for some relatively simple tasks, we'll go over some basics of how you might interact with data.

## Reading in files

One of the first steps in any type of programming analysis is dealing with the input and outputs of reading in and writing out files. When we create new variables, all the data needed to perform operations with those variables are avaliable locally. However, often one must read in a file, process its content, and then write out the results. Reading in a file takes various forms in R, but it all generally follows the same pattern! Mainly, all the files can be read into the R environment with a sterotyped `read` function whose required input is the filepath to the file in question. 

Some examples:

```
csv <- read.csv('filepath-to-csv-file')
txt <- read.table('filepath-to-.txt-file')
delim <- read.delim('filepath-to-delimited-file')
```

We will be dealing with clinical data from the METABRIC study for this exercise and reading in a table from a tab-delimited file. The first line of the file contains field headers, which are simply column names for the data in the file. These column names are important for interpreting the file contents, but we do not want to include them as values for our data. Consequently, we list the read command with **`header = TRUE`** which tells R to not count the first line as data. 

```{r cars}
clinical_df <- read.delim("/home/shared/data/metabric_split/activity_clinical.txt", header = TRUE)
```

Note that we've saved the data in a variable named `clinical_df`. The "df" in this case is short for "data frame", which is a kind of table that allows you to store different types of data together (as you'll see below).

## Inspecting the data

An important first step anytime one reads in a file is to inspect its contents and investigate the data types in each field/column. A nice way to quickly inspect the data is to simply visually inspect the contents of the file. For example if we desire to look at the first six lines of the file we can use the **`head()`** command.

```{r}
head(clinical_df)
```

Let's say we want to see more of the data. We can adjust the `head` command to show however many lines we want to see. For example let's look at the first 50 lines of data! In order to do this we simply need to provide a value for n, where n is an integer value representing the amount of lines from the top of the file you wish to see. 

If you were working in the command console, the following code would print out all 25 lines at once; a nice feature of R Notebooks is that it splits tables into multiple pages that you can scroll through.

```{r}
head(clinical_df, n = 25)
```

Just as we can inspect first n lines from the top of the file, we can likewise do the same for the bottom of the file. To view the contents at the end of a file we can use the **`tail()`** command. 

```{r}
tail(clinical_df)
```

Just like with the `head` command, we can adjust how many lines from the bottom of the file we wish to see by adjusting the n value.

```{r}
tail(clinical_df, n = 5)
```

## Checking data types

Visually inspecting the data suggests that several data types exist in the data frame. By means of example let's subset three fields in the data: survival time (`T`), `age_at_diagnosis`, `histological_type`, and survival. To subset the data we can use the **`$`** character which allows us to extract a particular field from the data frame. We will only look at the first 10 lines in each field, but you can subset the whole field by simply omitting the `head` command from the lines below.

```{r}
head(clinical_df$T)
head(clinical_df$age_at_diagnosis)
head(clinical_df$last_follow_up_status)
head(clinical_df$survival_5y)
```

Looking at the data from these four fields would suggest that our data includes at least 4 data types. Time (`T`) appears to be composed of integers (in this case, number of days), `age_at_diagnosis` appears to be composed of numerics (continuous numbers for fractions of years), `last_follow_up_status` appears to be composed of strings (letters indicating survival status), and `survival_5y` appears to be composed of boolean (indicating whether or not the patient survived at least 5 years) values. However, what's up with these "levels"? 

Let's take a closer look at the fields and specifically ask R to give us their data type with the `class()` function. 

```{r}
class(clinical_df$T)
class(clinical_df$age_at_diagnosis)
class(clinical_df$last_follow_up_status)
class(clinical_df$survival_5y)
```

When we specifically ask R to describe the data types, we find that while we correctly surmised that `T` contains integers and `age_at_diagnosis` contains numerics, it seems that we didn't get the class type right for either `last_follow_up_status` or `survival_5y`. 

These two fields were imported as **"factor"** values, which is R's way of representing catagorical variables. Factors can be helpful for when we begin modeling the data, but it can also be confusing if we think we are dealing with data types we are not. 

For example, logical values have some specific and useful functionality in R. Most often, they're used to represent a status or answer to a question. The next command asks the question "is `1` equal to `2`?" using the `==` symbols.

```{r}
1 == 2
```

Whereas this command asks "is `2` greater than `1`?"

```{r}
2 > 1
```

Logical values can be inverted with the `!` symbol. Check how it works with the next chunk.

```{r}
TRUE
!TRUE
```

However, if the values we were expecting to behave as logicals were in fact strings, this could lead to an error!

```{r}
"True"
!"True"
```

R's default behavior when reading in data is to treat strings (i.e., character values) as factors. This can lead to some unintended (and unexpected) consequences. One common practice is to avoid this behavior with the argument `stringsAsFactors = FALSE` when using any of the `read` functions. Alternatively, we might modify or correct data types after importing. This process is often referred to as data "cleaning" or "munging" — while it's not the most exciting part of data science, it can be extremely important. You'll see some more examples of this in subsequent modules.

## Summarizing data

Let's take a look at the **`summary`** command and it's usefulness. As you recall the `age_at_diagnosis` and the `T` data types were numerics and integers, respectively. If we run `summary()` on these data, we will literally get the summary statistics for them.

```{r}
summary(clinical_df$age_at_diagnosis)
summary(clinical_df$T)
```

For a string or factor field, `summary()` gives us the counts for each unique value in the data.

```{r}
summary(clinical_df$last_follow_up_status)
```


However, the power behind the `summary` function is made even more impressive when we apply it to the entire data frame.

```{r}
summary(clinical_df)
```

In one command we are able to summarize the data across the data frame!

---

# Submitting to the mini-DREAM Challenge

Below, we'll give a quick demonstration of what it's like to submit to the mini-DREAM challenge and get on the scoreboards.

## Synapse login

First, let's make sure that you can connect and log into Synapse. We'll do this in a way that the local installation of Synapse "remembers" you, and you won't need to enter your email and password each time.

Try to log in with the next chunk. If the chunk progress indicator stays *green* and seems to be hanging, you'll need to expand the **Console** window below to enter your credentials. After you finish that step, you can expand the `module1.Rmd` editor tab again to continue. If you get stuck at this point, just ask and we can help out!

**Note:** just to be safe, we'll also clear your R history, so that others can't see your Synapse login information.

```{r}
library(synapseClient)
synapseLogin(rememberMe = TRUE)
if (file.exists(".Rhistory")) {
    file.remove(".Rhistory")
}
```

If everything worked, you should see a "Welcome" message (if you also see the word "Goodbye", that's actually just a bug with the current version of the R Synapse package).

## Predicting survival time

For the mini-DREAM challenge submissions, we'll be using a separate, special dataset where *we don't know* the survival time (we do know, but it will be hidden in the activities). We've made an "R Data" file with this dataset to make it easier to load.

```{r}
load("/home/shared/data/metabric_split/challenge_data.RData", verbose = TRUE)
```

We'll be predicting survival time (in days) for samples in the challenge data. Let's see how many samples that is.

```{r}
nrow(challenge_clinical_df)
```

For this first submission, you'll actually just be making **random** predictions about patient survival. We'll create a new data frame that includes patient IDs along with predicted times (in *days*). To get these times, we'll randomly sample from a normal distribution (think bell curve), where the average survival time is 5 years (5x365 days), and the standard deviation is 1 year (365 days).

**Note:** the command `set.seed(1234)` is used to make sure these results are identical — or **reproducible — each time we run this code chunk.

```{r}
set.seed(1234)
prediction <- data.frame(METABRIC_ID = challenge_clinical_df$METABRIC_ID, 
                         T = rnorm(n = 434, mean = 5*365, sd = 365))
```

Let's check our predictions, to see if they make sense.

```{r}
summary(prediction$T)
```

None of the predictions are negative (i.e., the minimum is greater than 0) or impossibly long (the max, 2936, is about 8 years), so this looks good.

## Submitting the prediction

You're now ready to submit the prediction. Just run the chunk below, a file with your prediction will be uploaded to Synapse and submitted to the challenge. You'll be able to see the results of your prediction on the mini-DREAM scoreboards, with the submission ID that gets printed below.

```{r}
submission_filename <- paste(Sys.getenv("USER"), "activity-1.txt", sep = "_")
write.csv(prediction, submission_filename)
activity1_submission <- synStore(File(path = submission_filename, parentId = "syn9979188"))
submission <- submit(evaluation = "9604686", entity = activity1_submission)
print(submission[[1]]$id)
```

---

# Wrapping up

Congrats — you’ve reached the end of **Module 1**! You can now return to the **mini-DREAM Challenge** site on Synapse.


